Intentional release of native species undermines ecological stability

### 2022-10-14 rmd/manuscript.Rmd
rmd

---
title: "Intentional release of native species undermines ecological stability"
output:
  #bookdown::pdf_document2:
  #  latex_engine: xelatex
  bookdown::word_document2:
   reference_docx: word_format.docx
csl: https://www.zotero.org/styles/pnas
knit: (function(inputFile, encoding) {
        rmarkdown::render(inputFile,
                          encoding = encoding, 
                          output_dir = "document_output")
      })
bibliography: reference.bib
fontsize: 11pt
header-includes:
  - \usepackage{lineno}
  - \linenumbers
---

@@

Human demands for natural resources are ever-increasing, such that active interventions are critical to the sustainable management of fisheries, forestry, and wildlife [@laikreCompromisingGeneticDiversity2010]. Releases of captive-bred native species ("intentional release") is a form of the efforts to enhance wild populations of diverse taxa [@snyderLimitationsCaptiveBreeding1996; @laikreCompromisingGeneticDiversity2010]. The magnitude of intentional release for commercial and recreational species is massive; a single program releases tens of thousands to millions of plant [@laikreReleaseAlienPopulations2006], invertebrate (bivalves, insects) [@ingsBumblebeesHumblePollinators2005; @haagNorthAmericanFreshwater2012], and vertebrate individuals (fish, birds) [@villanuaSanitaryRisksRedlegged2008; @kitadaEconomicEcologicalGenetic2018] into the wild annually (reviewed in [@laikreCompromisingGeneticDiversity2010; @kitadaEconomicEcologicalGenetic2018]). There is growing awareness that intentional release entails ecological risks, including the accumulation of deleterious alleles [@arakiGeneticEffectsCaptive2007], the spread of diseases and parasites in wild populations [@krkosekDecliningWildSalmon2007], and the intensification of intraspecific competition within the enhanced species [@kitadaLessonsLearnedJapanese2006]. Yet, this method is still pervasive in natural resource management likely because of the apparent economic benefit of release programs. For example, Kitada @kitadaEconomicEcologicalGenetic2018 analyzed economic performance of release programs for 12 major fishery resources -- the gross monetary yield exceeded the seed-production cost in nine species, suggesting that release programs are economically profitable in general.

Current debates, however, overlook the fact that we have rarely assessed the community-wide impact of intentional release. Species are all embedded in the complex web of interacting organisms, and the stable coexistence of competing species through density-dependent feedback underpins the emergent stability of ecological communities [@tilmanEcologicalConsequencesChanges1999; @carrollNicheFitnessDifferences2011a; @loreauDoesFunctionalRedundancy2004]. Intentional release may disrupt the sensitive balance of species interactions because it introduces unnaturally high numbers of individuals into the wild [@laikreCompromisingGeneticDiversity2010; @careyNativeInvadersChallenges2012; @kitadaEconomicEcologicalGenetic2018]. Hence, this form of species management may intervene in the ecological process that allows competing species to coexist, ultimately degrading the long-term community stability. The community-wide impact must be particularly serious in release programs for fishery resources as many of them release millions of hatchery individuals into a single water body (e.g., a watershed) every year, which is likely far beyond the environmental capacity of the recipient ecosystem. Evidence for this hypothesis is lacking, however.

Here, we show that intentional release undermines long-term community stability, which we define as the relative size of fluctuations in total community density over time [@tilmanEcologicalConsequencesChanges1999]. Our theory illuminates that intentional release compromises the stabilizing mechanism emerging from species niche differences. The present study further demonstrates the relevance of this general theory to natural systems by showing its congruence with Japanese stream fish communities, where \~10 million hatchery masu salmon (*Oncorhynchus masou masou*) are released annually for fisheries, recreation, and conservation purposes [@kitadaLessonsJapanMarine2020].

>>>>

---
title: "Intentional release of native species undermines ecological stability"
output:
  bookdown::pdf_document2:
   latex_engine: xelatex
   toc: false
  #bookdown::word_document2:
  # reference_docx: word_format.docx
csl: https://www.zotero.org/styles/pnas
knit: (function(inputFile, encoding) {
        rmarkdown::render(inputFile,
                          encoding = encoding, 
                          output_dir = "document_output")
      })
bibliography: reference.bib
fontsize: 11pt
mainfont: "Times New Roman"
header-includes:
  - \usepackage{lineno}
  - \linenumbers
---

@@

Human demands for natural resources are ever-increasing, such that active interventions are critical to the sustainable management of fisheries, forestry, and wildlife [@laikreCompromisingGeneticDiversity2010]. Releases of captive-bred native species ("intentional release") is a form of the efforts to enhance wild populations of diverse taxa [@snyderLimitationsCaptiveBreeding1996; @laikreCompromisingGeneticDiversity2010]. In particular, the magnitude of intentional release for commercial and recreational species is massive (reviewed in [@laikreCompromisingGeneticDiversity2010; @kitadaEconomicEcologicalGenetic2018]) -- a single program releases up to tens of millions of plant [@laikreReleaseAlienPopulations2006], invertebrate (bivalves, insects) [@ingsBumblebeesHumblePollinators2005; @haagNorthAmericanFreshwater2012], and vertebrate individuals (fish, birds) [@villanuaSanitaryRisksRedlegged2008; @kitadaEconomicEcologicalGenetic2018] into the wild annually. There is growing awareness that intentional release entails ecological risks, including the accumulation of deleterious alleles [@arakiGeneticEffectsCaptive2007], the spread of diseases in wild populations [@krkosekDecliningWildSalmon2007], and the intensification of intraspecific competition within the enhanced species [@kitadaLessonsLearnedJapanese2006]. Yet, this method is still pervasive in natural resource management owing to the significant economic benefit [@malkamakiSystematicReviewSocioeconomic2018; @davidEconomicFeasibilityIntensification2018; @rutledgeCostBenefitHatcheryreared1990; @kitadaEconomicEcologicalGenetic2018]. For example, Kitada @kitadaEconomicEcologicalGenetic2018 analyzed economic performance of release programs for 12 major fishery resources across the globe. The gross monetary yield exceeded the seed-production cost in nine species, suggesting that release programs are economically profitable in general.

Current debates, however, overlook the fact that we have rarely assessed the community-wide impact of intentional release. Species are all embedded in the complex web of interacting organisms, and the stable coexistence of competing species through density-dependent feedback underpins the emergent stability of ecological communities [@tilmanEcologicalConsequencesChanges1999; @carrollNicheFitnessDifferences2011a; @loreauDoesFunctionalRedundancy2004]. Intentional release may disrupt the sensitive balance of species interactions because it introduces unnaturally high numbers of individuals into the wild [@laikreCompromisingGeneticDiversity2010; @careyNativeInvadersChallenges2012; @kitadaEconomicEcologicalGenetic2018]. Hence, this form of species management may intervene in the ecological process that allows competing species to coexist, ultimately degrading the long-term community stability. The community-wide impact may be particularly serious in release programs for fishery resources as many of them release millions of hatchery individuals into a single water body (e.g., a watershed) every year. Evidence for this hypothesis is lacking, however.

Here, we show that intentional release undermines long-term community stability, which we define as the relative size of fluctuations in total community density over time [@tilmanEcologicalConsequencesChanges1999]. Our theory illuminates that intentional release compromises the stabilizing mechanism emerging from species niche differences. The present study further demonstrates the relevance of this general theory to natural systems by showing its congruence with Japanese stream fish communities, where \~10 million hatchery masu salmon (*Oncorhynchus masou masou*) are released annually for fisheries, recreation, and conservation purposes across the nation [@kitadaLessonsJapanMarine2020]. Once released as fry, masu salmon stay in freshwater for at least one year before sea migration and compete for resources with other stream-dwelling species. Therefore, this system serves as an excellent model to study the community-wide impact of intentional release. Our integrative approach provides strong evidence for the destabilizing effect of intentional release on ecological communities.

### 2022-10-14 rmd/supporting_info.Rmd
rmd

The analysis of a two-species community revealed the distinct parameter space predicting a stabilizing, neutral, or destabilizing effect of intentional release. The left column in **Figure S**\@ref(fig:fig-2sp-w)**A**) shows the predictions in scenarios with no environmental stochasticity ($\sigma_{\varepsilon} =$ `r min(list_param[[1]]$sd_env)`) and weak competition ($\bar{\alpha}=$ `r min(list_param[[1]]$alpha)`). For lower values of intrinsic growth rate $r_i$ ($<1.5$), intentional release had neutral (with higher $K$) to destabilizing effects (with lower $K$) on community dynamics. In the meantime, stabilization with intentional release occurred for higher values of $r_i$ ($\ge 1.5$) by dampening limit cycles or chaos (**Figure S**\@ref(fig:fig-2sp-s)), which is reflected in decreased SD $\sigma$ (**Figure S**\@ref(fig:fig-2sp-w)). However, the long-term mean of community density $\mu$ consistently decreased with intentional release in almost all cases except for only a few combinations of $r_i$ and $K$ (**Figure S**\@ref(fig:fig-2sp-w)).

>>>>

The analysis of a two-species community revealed the distinct parameter space predicting a stabilizing, neutral, or destabilizing effect of intentional release. The left column in **Figure S**\@ref(fig:fig-2sp-w)**A** shows the predictions in scenarios with no environmental stochasticity ($\sigma_{\varepsilon} =$ `r min(list_param[[1]]$sd_env)`) and weak competition ($\bar{\alpha}=$ `r min(list_param[[1]]$alpha)`). For lower values of intrinsic growth rate $r_i$ ($<1.5$), intentional release had neutral (with higher $K$) to destabilizing effects (with lower $K$) on community dynamics. In the meantime, stabilization with intentional release occurred for higher values of $r_i$ ($\ge 1.5$) by dampening limit cycles or chaos (**Figure S**\@ref(fig:fig-2sp-s)), which is reflected in decreased SD $\sigma$ (**Figure S**\@ref(fig:fig-2sp-w)). However, the long-term mean of community density $\mu$ consistently decreased with intentional release in almost all cases except for only a few combinations of $r_i$ and $K$ (**Figure S**\@ref(fig:fig-2sp-w)).
